---
title: "Exploring wellbeing: How does education and income impact individual wellbeing?"
author: '660032036'
date: "18 April 2018"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Health is an enormously important topic in sociology; for example, unequal morbidity and mortality rates are key indicators of social inequalities. However, while class differences have been shown to be one of the “most enduring and incontrovertible indicators” of perceptions of personal health (referred to below as wellbeing), it has been argued that more research is required in order to fully understand the dynamics of the social divides in wellbeing (Bennet et. al., 2009: 152; Mackenbach, 2012; Pinxten and Lievens, 2014). This essay aims to further current research into wellbeing by analysing the impact of sex, education and income. In the UK, education and waged-work  are central to most people’s lives, thus likely to be near-universal influences on wellbeing, while it is also important to analyse the effects of sex-differences, given that male suicide rates are over three times that of female suicide rates (Samaritans, 2017).

Our research questions and hypotheses are presented in *Table 1*.

Research Question  | Hypothesis
----------------------------------------- | -----------------------------------------
How is wellbeing distributed across educational qualifications, and is this relationship consistent over time? | The more educated an individual, the higher their reported wellbeing. This relationship is consistent over time.
How is wellbeing distributed across the sexes, and is this relationship consistent over time? | Women report higher wellbeing than men. This relationship is consistent over time.
What is the relationship between wellbeing and income, and does an individual's wellbeing change as their income changes? | There is a positive relationship between income and wellbeing. As individual income increases, so does individual wellbeing.

To carry out this analysis, we will be using the *Understanding Society* (University of Essex, 2017) dataset, a longitudinal study of a nationally representative sample of approximately 40,000 households in the United Kingdom, carried out from 2009-2016. Each participant aged 16 and over completed an individual, face-to-face interview and a self-completed questionnaire, with one person in each household completing the household questionnaire.

Over this time period, a number of national events could have affected wellbeing. The UK emerging from an economic recession, two general elections and the 2012 Olympics were all significant events that should be considered when analysing our data.

# Preparing the data

Our first task was to read the data into R.

```{r loading, cache = TRUE, message = FALSE, warning = FALSE, results = 'hide'}
# We first installed the necessary packages.

library(tidyverse)

library(data.table)

files <- dir("data/UKDA-6614-tab/tab",
             pattern="indresp",
             recursive = TRUE,
             full.names=TRUE)

files <- files[str_detect(files, "us")]

varsI <- c('pidp', 'qfhigh_dv', 'fimnnet_dv', 'fimngrs_dv', 'sclfsat1', 'dvage', 'sex', 'racel', 'racelo_code')

for (i in 1:7) {
  varsToSelectI <- paste(letters[i], varsI, sep = "_")
  varsToSelectI <- c("pidp", varsToSelectI)
  dataI <- fread(files[i], select = varsToSelectI)
  if (i == 1) {
    dI <- dataI  
  }
  else {
    dI <- full_join(dI, dataI, by = "pidp")
  }
  rm(dataI)
}
```

To arrange the data to facilitate analysis, the ‘wave’ the data was collected in needed to be recoded as a variable.

```{r join, cache = TRUE, message = FALSE, warning = FALSE}
dmeltI <- dI %>%
  melt(id = "pidp") %>%
  separate(variable, into = c("wave", "variable"), sep = "_") %>%
  dcast(pidp + wave ~ variable)
```

The next step in preparing our data was removing the values for each variable that represented, for example, respondents refusing to answer the question, and changing these to ‘NA’ values.

```{r NA, cache = TRUE, message = FALSE, warning = FALSE}
dnaI <- dmeltI %>%
  mutate(qfhigh = ifelse(qfhigh > 0, qfhigh, NA)) %>%
  mutate(qfhigh = ifelse(qfhigh < 96, qfhigh, NA)) %>%
  mutate(sclfsat1 = ifelse(sclfsat1 > 0, sclfsat1, NA)) %>%
  mutate(dvage = ifelse(dvage > 0, dvage, NA)) %>%
  mutate(sex = ifelse(sex > 0, sex, NA)) %>%
  mutate(racel = ifelse(racel > 0, racel, NA)) %>%
  mutate(racel = ifelse(racel < 96, racel, NA)) %>%
  mutate(racelo = ifelse(racelo > 0, racelo, NA)) %>%
  mutate(racelo = ifelse(racelo < 96, racelo, NA))
```

Initially, we aimed to include ethnicity in our analysis; a factor Grineski (2009) identifies as influencing wellbeing. However, the variables for ethnicity had vast numbers of ‘NA’ values in each wave, with the ethnicity of the participant coded inconsistently across the waves of the data. Due to the limitations of this assignment, we decided simply to remove this variable.

```{r race, cache = TRUE, message = FALSE, warning = FALSE}
# Subsetting the data frame into individual waves
W1 <- subset(dnaI, wave == 'a', select = c(1:10))

W7 <- subset(dnaI, wave == 'g', select = c(1:10))

# Checking for variables with larger than average number of NAs
summary(W1)

summary(W7)

dnaI <- subset(dnaI, , -c(racel, racelo, fimngrs))

# We have also removed the 'gross monthly income' variable, as working with 'net monthly income' should allow a sufficiently clear look at income and wellbeing.
```

The final step in preparing our data involved, firstly, recoding the answers coded as numerical values to descriptive labels. We also combined equivalent qualifications, thereby reducing the number of variables. These changes made our descriptive analyses easier to carry out and the results easier to interpret.

```{r clean, cache = TRUE, message = FALSE, warning = FALSE}
dcleanI <- na.omit(dnaI) %>%
  mutate(qfhigh = recode(qfhigh, "1" = "1-Higher degree",
                         "2" = "2-Degree/equiv",
                         "3" = "3-Higher ed",
                         "4" = "3-Higher ed",
                         "5" = "3-Higher ed",
                         "6" = "3-Higher ed",
                         "7" = "4-A level/equiv",
                         "8" = "4-A level/equiv",
                         "9" = "4-A level/equiv",
                         "10" = "5-AS level/equiv",
                         "11" = "5-AS level/equiv",
                         "12" = "5-AS level/equiv",
                         "13" = "6-GCSE/equiv",
                         "14" = "7-CSE",
                         "15" = "6-GCSE/equiv",
                         "16" = "8-Other school cert")) %>%
  mutate(qfhigh = factor(qfhigh)) %>%
  mutate(sclfsat1 = recode(sclfsat1, "1" = "1-completely dissatisfied",
                           "2" = "2-mostly dissatisfied",
                           "3" = "3-somewhat dissatisfied",
                           "4" = "4-neither satisfied or dissatisfied",
                           "5" = "5-somewhat satisfied",
                           "6" = "6-mostly satisfied",
                           "7" = "7-completely satisfied")) %>%
  mutate(sclfsat1 = factor(sclfsat1)) %>%
  mutate(sex = recode(sex, "1" = "male", "2" = "female")) %>%
  mutate(sex = factor(sex)) %>% 
  mutate(year = recode(wave, "a" = "2009",
                       "b" = "2010",
                       "c" = "2011",
                       "d" = "2012",
                       "e" = "2013",
                       "f" = "2014",
                       "g" = "2015")) %>%
  mutate(year = as.numeric(year))
```

Secondly, we recorded the initial responses to wellbeing as numerical values, rather than the labels used above, in order to treat wellbeing as a continuous variable and carry out statistical analysis of the effects of our variables on it.

```{r cont, cache = TRUE, message = FALSE, warning = FALSE}
dcontI <- na.omit(dnaI) %>%
  mutate(qfhigh = recode(qfhigh, "1" = "1-Higher degree",
                         "2" = "2-Degree/equiv",
                         "3" = "3-Higher ed",
                         "4" = "3-Higher ed",
                         "5" = "3-Higher ed",
                         "6" = "3-Higher ed",
                         "7" = "4-A level/equiv",
                         "8" = "4-A level/equiv",
                         "9" = "4-A level/equiv",
                         "10" = "5-AS level/equiv",
                         "11" = "5-AS level/equiv",
                         "12" = "5-AS level/equiv",
                         "13" = "6-GCSE/equiv",
                         "14" = "7-CSE",
                         "15" = "6-GCSE/equiv",
                         "16" = "8-Other school cert")) %>%
  mutate(qfhigh = factor(qfhigh)) %>%
  mutate(sex = recode(sex, "1" = "male", "2" = "female")) %>%
  mutate(sex = factor(sex)) %>% 
  mutate(year = recode(wave, "a" = "2009",
                       "b" = "2010",
                       "c" = "2011",
                       "d" = "2012",
                       "e" = "2013",
                       "f" = "2014",
                       "g" = "2015")) %>%
  mutate(year = as.numeric(year))
```

Finally, before visualising the relationships between our variables and wellbeing, we removed the data files we no longer needed to save memory.

```{r rm, cache = TRUE, message = FALSE, warning = FALSE}
rm(dI, dmeltI, dnaI)
```

# Descriptive analysis

## Distribution of wellbeing

First, we looked for any overall trends in the distribution of wellbeing across each wave of the data. We computed and visualised the mean average wellbeing score in each wave, which showed a noticeable drop in average wellbeing from 2010 to 2011, which (despite a small increase in 2012) remained relatively unchanged until increasing again in 2014 (*Figure 1*).

```{r wellbeing by year, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
  group_by(year) %>%
  summarise(
   meanWb = mean(sclfsat1, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = year, y = meanWb)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 1",
       subtitle = "Mean Wellbeing by Year",
       x = "Year",
       y = "Mean Wellbeing")
```

For a more detailed look at the trend in wellbeing over time, we looked at the proportion of responses for each level of wellbeing in each year of our data (*Figure 2*). The proportion of people both ‘mostly’ and ‘completely’ dissatisfied with their wellbeing increased dramatically from 2010 to 2011, whilst the proportion of people ‘mostly satisfied’ fell. Interestingly, over the years 2011-2013 (the period with the lowest mean wellbeing), though the proportion of people either ‘mostly’ or ‘completely’ dissatisfied with their wellbeing slowly decreased, so did the proportion ‘completely satisfied’ with their wellbeing. The proportion of people ‘mostly’ and ‘completely’ satisfied with their wellbeing increased 2014 and continued to rise in 2015, with those ‘mostly’ and ‘completely’ dissatisfied eventually returning to their 2009 proportion.

```{r wellbeing breakdown, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(dcleanI, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 2",
       subtitle = "Proportion of Wellbeing responses by Year",
       x = "Year",
       y = "Proportion of Wellbeing responses")

# Note: This legend applies for all other charts with this colour scheme, unless otherwise stated. To make future charts easier to interpret, refer back to this legend.
```

*Figure 3* allows us a closer look at the contrast in wellbeing across the first, fourth and final waves of the data (‘a’, ‘d’ and ‘g’ respectively).

```{r select few, cache = TRUE, message = FALSE, warning = FALSE}
Wc <- subset(dcleanI, wave == c('a', 'd', 'g'), select = c(1:8))

ggplot(Wc, aes(x = wave, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 3",
       subtitle = "Proportion of Wellbeing responses in Waves 1, 4 & 7",
       x = "Wave",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")
```

## Differentiating by sex

In order to examine more closely the distribution of wellbeing across society, we first differentiated our data by sex. This allowed us to perform our statistical analysis while controlling for sex, which, as per our hypothesis, is expected to have an effect on wellbeing. First, we created a simple line graph, showing the mean average wellbeing score for each wave (*Figure 4*).

```{r sex time, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
  group_by(year, sex) %>%
  summarise(
    meanWb = mean(sclfsat1, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = sex, linetype = sex, shape = sex)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 4",
       subtitle = "Mean Wellbeing split by Sex",
       x = "Year",
       y = "Mean Wellbeing")
```

This shows a clear and consistent difference in wellbeing between the sexes, as per our hypothesis. Interestingly, however, average wellbeing for men is higher than for women (though only slightly), going against our hypothesis that men will have a lower average wellbeing than women.

Next, for an overview of the differences between wellbeing of men and women over the entire dataset, we created a coxcomb-chart, split by sex (*Figure 5*). This shows that, overall, women are proportionally more likely to be ‘completely dissatisfied’ with their wellbeing than men, though have comparable levels of being ‘completely satisfied’.

```{r sex coxcomb, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(dcleanI, aes(x = sex, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  coord_polar(theta = 'y') +
  labs(title = "Figure 5",
       subtitle = "Proportion of Wellbeing responses split by Sex",
       x = "Sex",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")
```

To closer examine this difference, we compared the individual responses to wellbeing given by respondents over each wave. *Figures 6* & *7* show that for 2009, 2010 and 2014, the proportion of women who were ‘completely satisfied’ with their wellbeing is higher than for men. However, the reverse is true in every positive category apart from these three. For each year, there was a higher proportion of women who were ‘completely’ and ‘mostly’ dissatisfied with their wellbeing than men.

```{r sex split, cache = TRUE, message = FALSE, warning = FALSE}
fe1 <- subset(dcleanI, sex == 'female')

m1 <- subset(dcleanI, sex == 'male')

ggplot(fe1, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 6",
       subtitle = "Proportion of Wellbeing responses for Women",
       x = "Year",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")

ggplot(m1, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 7",
       subtitle = "Proportion of Wellbeing responses for Men",
       x = "Year",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")
```

Another useful insight from this analysis is how wellbeing has changed for men and women over time, in relation to one another. Returning to *Figure 4*, we can see that while the overall trend is similar (a dip in wellbeing over 2011-2013), the difference in mean wellbeing between men and women is not constant. Whether this difference is statistically significant is analysed below.

## Differentiating by education

We continued our analysis by differentiating wellbeing by education. As  in our analysis of sex and wellbeing, we first made a simple line graph showing the mean average wellbeing score for each wave, segregated by level of education (*Figure 8*).

```{r edu, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
 group_by(year, qfhigh) %>%
 summarise(
   meanWb = mean(sclfsat1, na.rm = TRUE)
 ) %>%
 ggplot(aes(x = year, y = meanWb, colour = qfhigh)) +
 geom_point() +
 geom_line() +
  labs(title = "Figure 8",
       subtitle = "Mean Wellbeing split by Education",
       x = "Year",
       y = "Mean Wellbeing")
```

This shows some clear differences between levels of education and wellbeing, though not entirely as predicted. A- and AS-levels appear to be interchangeable when evaluating their effect on wellbeing, although in 2013 those with AS-levels had, proportionally, an unusually high mean average wellbeing (even higher than those with a degree or equivalent). Further, those with AS-levels were the only group whose wellbeing did not increase from 2011 to 2012. Interestingly, respondents with higher education (e.g. 'teaching qualification (excluding PGCE)’) have similar scores of wellbeing as those with just GCSE/equivalent qualifications; noticeably lower than those with AS- or A-levels, or university degrees. Finally, those with ‘Other school certificates’ have a higher mean average than those with ‘CSEs', with the exception of 2015.

For a more detailed examination, we created a stacked bar-chart showing the overall proportions of wellbeing for each education level (*Figure 9*). Interestingly, this shows that respondents with AS- and A-levels are more likely to be ‘completely satisfied’ with their wellbeing than those with university degrees or higher, a finding that was not picked up in the previous chart. For a clearer comparison, we added a coxcomb chart comparing respondents with higher degrees, A-levels, and CSEs (*Figure 10*).

```{r edu breakdown, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(dcleanI, aes(x = qfhigh, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 9",
       subtitle = "Proportion of Wellbeing responses split by Education",
       x = "Education",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")

Wd <- subset(dcleanI, qfhigh == c('1-Higher degree',
                                  '4-A level/equiv',
                                  '7-CSE'), select = c(1:7))

ggplot(Wd, aes(x = qfhigh, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  coord_polar(theta = 'y') +
  labs(title = "Figure 10",
       subtitle = "Proportion of Wellbeing responses split for 'CSEs', 'A-Levels' and 'Higher degrees'",
       x = "Education",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")
```

While this analysis is more detailed than our initial line-graph, we were able to develop this further by analysing these three qualification levels by each wave of data. *Figures 11*, *12*, and *13* show us that during the overall slump in wellbeing from 2011-2013, the proportion of those who were ‘completely’ and ‘mostly’ dissatisfied with their wellbeing increased the most in respondents with the highest education levels. However, for those ‘completely’ and ‘mostly’ satisfied with their wellbeing, it is those with A-levels, not with higher degrees who were least affected; in fact, the proportion of those with A-levels who were ‘completely satisfied’ with their wellbeing actually increased from 2010-2011.

```{r edu few, cache = TRUE, message = FALSE, warning = FALSE}
High <- subset(dcleanI, qfhigh == '1-Higher degree', select = c(year, sclfsat1))

A.l <- subset(dcleanI, qfhigh == '4-A level/equiv', select = c(year, sclfsat1))

CSE <- subset(dcleanI, qfhigh == '7-CSE', select = c(year, sclfsat1))

ggplot(High, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 11",
       subtitle = "Proportion of Wellbeing responses for 'Higher degrees'",
       x = "Year",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")

ggplot(A.l, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 12",
       subtitle = "Proportion of Wellbeing responses for 'A-Levels'",
       x = "Year",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")

ggplot(CSE, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 13",
       subtitle = "Proportion of Wellbeing responses for 'CSEs'",
       x = "Year",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")
```

Following our analysis of the split between the sexes and wellbeing, it would be remiss not to include sex when analysing education. *Figure 14* shows that the split between the sexes is exaggerated by certain education levels. Those who are least well educated (CSEs) have substantially higher levels of wellbeing for men than for women. Men with GCSEs, A-levels and higher education report similar levels of wellbeing, whereas women with GCSEs and higher education have substantially lower levels of wellbeing than those with A-levels. Interestingly, men with AS-levels and higher university degrees have the highest levels of wellbeing in men, whereas women with higher university degrees and university degrees have the highest levels of wellbeing in women. The only qualification where women score higher on wellbeing than men is university degrees.

```{r edu sex, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
 group_by(year, qfhigh, sex) %>%
 summarise(
   meanWb = mean(sclfsat1, na.rm = TRUE)
 ) %>%
 ggplot(aes(x = year, y = meanWb, colour = qfhigh, linetype = sex, shape = sex)) +
 geom_point() +
 geom_line() +
  labs(title = "Figure 14",
       subtitle = "Mean Wellbeing split by Education and Sex",
       x = "Year",
       y = "Mean Wellbeing")
```

## Differentiating by income

The final part of our descriptive analysis brought us on to the topic of income. We recoded net monthly income into quintiles to make our descriptive analysis easier to do and to interpret.

```{r income code, cache = TRUE, message = FALSE, warning = FALSE}
quantile(dcleanI$fimnnet, seq(0, 1, 0.2))

sc <- dcleanI %>% 
  mutate(income = ifelse(fimnnet >= -17741.3379 & fimnnet <= 518.9808, 20,
                         ifelse(fimnnet > 518.9808 & fimnnet <= 1051.6666, 40,
                                ifelse(fimnnet > 1051.6666 & fimnnet <= 1502.5000, 60,
                                       ifelse(fimnnet > 1502.5000 & fimnnet <= 2150.7295, 80,
                                              ifelse(fimnnet > 2150.7295 & fimnnet <= 15000.0000, 100, NA)))))) %>% 
  mutate(income = as.factor(income))
```

*Figure 15* is an overview of the dataset when segregating respondents by income. Incredibly, the group who are proportionally most likely to be ‘completely satisfied’ with their wellbeing are the bottom 20% of earners. More precisely, tabulating the raw numbers shows that the bottom 20% of earners are the third least likely to be ‘completely dissatisfied’, second least likely to be ‘mostly dissatisfied’ and least likely to be ‘somewhat dissatisfied’ in terms of wellbeing. These factors, coupled with the fact that the 4 other income brackets have almost identical proportions of respondents who were ‘completely satisfied’ with their wellbeing, is entirely contrary to our hypothesis that income is positively correlated with wellbeing.

```{r income overview, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(sc, aes(x = income, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  labs(title = "Figure 15",
       subtitle = "Proportion of Wellbeing responses for Income",
       x = "Income Quintile",
       y = "Proportion of Wellbeing responses") +
  theme(legend.position = "none")

table(sc$sclfsat1, sc$income)
```

To evaluate this further, we created a line-graph of income and wellbeing. *Figure 16* shows almost exactly what we hypothesised: as income decreases, so does the mean average wellbeing. However, there is one substantial exception to this rule. Those in the bottom fifth of earners have exceptionally high mean average wellbeing levels, not far from the top fifth of earners. This very surprising result could, however, be quite easy to explain; net monthly income can be negative. As this variable is purely personal income, it does not take into account other earners in the household. The household net income may be in the top 20%, yet earnt by only one person in the household, meaning other family members benefit from this secondary wealth. To try to counter this possible effect, we removed individuals who had a negative net income, as we hypothesised that those on low-salaries would not spend far outside their means, unlike those with a spouse or parent on a high-income salary (*Figure 17*).

```{r income t & s, cache = TRUE, message = FALSE, warning = FALSE}
# We first had to recode the data into quintiles using our data with wellbeing coded as a continuous variable.

quantile(dcontI$fimnnet, seq(0, 1, 0.2))

t <- dcontI %>% 
  mutate(income = ifelse(fimnnet >= -17741.3379 & fimnnet <= 518.9808, 20,
                         ifelse(fimnnet > 518.9808 & fimnnet <= 1051.6666, 40,
                                ifelse(fimnnet > 1051.6666 & fimnnet <= 1502.5000, 60,
                                       ifelse(fimnnet > 1502.5000 & fimnnet <= 2150.7295, 80,
                                              ifelse(fimnnet > 2150.7295 & fimnnet <= 15000.0000, 100, NA)))))) %>% 
  mutate(income = as.factor(income))

t %>%
  group_by(year, income) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 16",
       subtitle = "Mean Wellbeing split by Income",
       x = "Year",
       y = "Mean Wellbeing")

quantile(dcontI$fimnnet[dcontI$fimnnet > 0], seq(0, 1, 0.2))

s <- dcontI %>% 
  filter(fimnnet > 0) %>% 
  mutate(income = ifelse(fimnnet >= 1.506607e-04 & fimnnet <= 6.706440e+02, 20,
                         ifelse(fimnnet > 6.706440e+02 & fimnnet <= 1.125058e+03, 40,
                                ifelse(fimnnet > 1.125058e+03 & fimnnet <= 1.567589e+03, 60,
                                       ifelse(fimnnet > 1.567589e+03 & fimnnet <= 2.200000e+03, 80,
                                              ifelse(fimnnet > 2.200000e+03 & fimnnet <= 1.500000e+04, 100, NA)))))) %>% 
  mutate(income = as.factor(income))

s %>%
  group_by(year, income) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 17",
       subtitle = "Mean Wellbeing split by Income (of over £0p/m)",
       x = "Year",
       y = "Mean Wellbeing")
```

This change in coding had a substantial effect, and although the bottom fifth of earners still represent an outlier in mean average wellbeing, they are now closer to the second fifth of earners than the top fifth. Interestingly, they seemed to be least affected by the general slump in average wellbeing, going from second bottom to second top (in mean average wellbeing) from 2010 to 2013. One possible explanation for the bottom fifth still being an outlier is retirees, as though they may have relatively low-income, they are more likely to have savings available to spend. We correct for this by sub-setting our data to only include those under 65 (the retirement age in the UK). However, *Figure 18* shows that this has very little difference overall, suggesting other factors are present.

```{r income y1, cache = TRUE, message = FALSE, warning = FALSE}
y1 <- subset(s, dvage < 65, select = c(1:9))

table(y1$dvage)

y1 %>%
  group_by(year, income) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 18",
       subtitle = "Mean Wellbeing split by Income (of over £0p/m, excluding pensioners)",
       x = "Year",
       y = "Mean Wellbeing")
```

The final part of our descriptive analysis involves dividing income levels by sex, given the notable difference it made in wellbeing generally. Indeed, by segregating between sexes, *Figure 19* uncovers some remarkable trends. Similar to education, there is only one level of income at which women have a higher mean average wellbeing than men, in this case the second fifth of earners. For every other level of income, men have a higher mean average wellbeing each year, almost without exception. Just as noteworthy is the mean average wellbeing for the bottom fifth of earners when split by sex. For men, this group’s wellbeing hovers around the same level as those in the second highest bracket of earners, and nearly matches those in the top bracket of earners in the most recent wave of survey data. For women, the bottom 20% of earners report either the lowest, or close to the lowest, mean average wellbeing. Furthermore, this category has the greatest difference between male and female levels for wellbeing, with 2014 being the only notable exception.

```{r income y1 sex, cache = TRUE, message = FALSE, warning = FALSE}
y1 %>%
  group_by(year, income, sex) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income, linetype = sex, shape = sex)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 19",
       subtitle = "Mean Wellbeing split by Income (of over £0p/m, excluding pensioners) and Sex",
       x = "Year",
       y = "Mean Wellbeing")
```

Unfortunately, visualising the differences in wellbeing between respondents with different levels of income, education, and of different sexes is impractical, and this comparison will be left for statistical analysis (*Figure 20* is the graph split by each of these three factors. Aside from a couple of notable outliers, it is almost entirely uninterpretable).

```{r y1 all, cache = TRUE, message = FALSE, warning = FALSE}
y1 %>%
  group_by(year, income, sex, qfhigh) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income, linetype = sex, shape = qfhigh)) +
  geom_point() +
  geom_line() +
  labs(title = "Figure 20",
       subtitle = "Mean Wellbeing split by Income (of over £0p/m, excluding pensioners), Sex and Education",
       x = "Year",
       y = "Mean Wellbeing")
```

# Statistical analysis

## Analysing the most recent wave of data

Having taken a preliminary look into our dataset, and visualised some of the relationships of interest, we move on to evaluating whether any of these relationships are statistically significant. First, we compared the relationship between income and wellbeing for 2015, using a linear model to see if the relationship was statistically significant. We then visualised this relationship.

```{r income lm & graph, cache = TRUE, message = FALSE, warning = FALSE}
# Subsetting our data to just the final wave (2015).

W7 <- subset(dcontI, wave == 'g', select = c(1:8))

Model.1 <- lm(sclfsat1 ~ fimnnet, W7)

summary(Model.1)

W7 %>%
  ggplot(aes(x = fimnnet, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  labs(title = "Figure 21",
       subtitle = "Trend in Wellbeing by Income",
       x = "Net Monthly Income",
       y = "Wellbeing")
```

*Model 1* shows a statistically significant relationship (at the 99% confidence level, p<0.01) between income and wellbeing, though *Figure 21* shows that a linear model may not be best suited to capturing this relationship. As indicated in our descriptive analysis, the relationship between income and wellbeing is less clear for respondents with low, or negative, monthly income, for whom there seem to be a number of confounding factors. In fact, *Figure 22* shows a linear model is more suitable when excluding respondents who earn under £800 a month, equating to £9,600 a year. *Model 2* shows that following this change, the effect of income on wellbeing almost doubles (from 3.951e-05 to 7.256e-05 per pound). Interestingly, this is roughly the amount an individual on minimum wage, working a 40-hour week for 45 weeks (taking into account bank-holidays and standard holiday allowance) will earn after tax (shown below). It seems, therefore, that when in full time employment, wellbeing increases with income (until about £8,000 a month).

```{r minimum income, cache = TRUE, message = FALSE, warning = FALSE}
W7r1 <- subset(W7, !(fimnnet < 800), select = c(1:8))

W7r1 %>%
  ggplot(aes(x = fimnnet, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  labs(title = "Figure 22",
       subtitle = "Trend in Wellbeing by Income (above £800)",
       x = "Net Monthly Income",
       y = "Wellbeing")

Model.2 <- lm(sclfsat1 ~ fimnnet, W7r1)
summary(Model.2)

6.7*40*45*.8

# Calculated using the minimum wage for over-21s (https://www.gov.uk/national-minimum-wage-rates) for 2015 (the final year in which survey data was taken), subtracting holiday time (https://www.gov.uk/bank-holidays and https://www.gov.uk/holiday-entitlement-rights), after tax (https://www.gov.uk/income-tax-rates/previous-tax-years).
```

Next, we examined the relationship between education and wellbeing. *Model 3* showed that, contrary to our hypothesis, when compared to respondents with higher degrees, those with AS-levels did not have a statistically significant different wellbeing (at the 95% confidence level, p>0.05). However, the differences in wellbeing between respondents with higher degrees and all other qualifications *were* statistically significant (at the 99% confidence level, p<0.01). Interestingly, the effect on wellbeing for respondents who had gone through higher education (e.g. a teaching qualification but not a PGCE) and those with GCSEs was similar (-0.24682 and -0.22977 respectively), though the effect of A-levels was smaller (only -0.10773). The largest effects came from respondents with CSEs and ‘Other school certificates’ (-0.41846 and -0.52835 respectively). The wellbeing scores for these education levels are included in *Figure 23*.

```{r education, cache = TRUE, message = FALSE, warning = FALSE}
Model.3 <- lm(sclfsat1 ~ factor(qfhigh), W7)

summary(Model.3)

W7 %>%
  ggplot(aes(x = qfhigh, y = sclfsat1, fill = qfhigh)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  labs(title = "Figure 23",
       subtitle = "Wellbeing by Education",
       x = "Education",
       y = "Wellbeing") +
  theme(legend.position = "none")
```

Penultimately, we evaluated the relationship between wellbeing and sex. As our descriptive analysis suggests, men have a higher wellbeing score than women. *Model 4* allows us to see that the difference between the two is statistically significant at the 99% confidence interval (p<0.01). However, though significant, the magnitude of this effect is fairly small (0.10) and is visualised in *Figure 24*.

```{r sex model & graph, cache = TRUE, message = FALSE, warning = FALSE}
Model.4 <- lm(sclfsat1 ~ factor(sex), W7)

summary(Model.4)

W7 %>%
  ggplot(aes(x = sex, y = sclfsat1, fill = sex)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  labs(title = "Figure 24",
       subtitle = "Wellbeing by Sex",
       x = "Sex",
       y = "Wellbeing")
```

Finally, before creating a model that controls for all our variables, we calculated whether there was a significant relationship between the age of respondents and their wellbeing, having removed those over 64 from our descriptive analysis of income.

```{r age lm, cache = TRUE, message = FALSE, warning = FALSE}
Model.5 <- lm(sclfsat1 ~ dvage, W7)

summary(Model.5)
```

*Model 5* shows that, in our sample, the older the respondent, the lower reported wellbeing. This is statistically significant at the 99% confidence level (p<0.01). *Figure 25* gives a visual representation of this relationship.

```{r age graph, cache = TRUE, message = FALSE, warning = FALSE}
W7 %>%
  ggplot(aes(x = dvage, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  labs(title = "Figure 25",
       subtitle = "Trend in Wellbeing by Age",
       x = "Age",
       y = "Wellbeing")
```

This shows that a linear model seems to fit the relationship between age and wellbeing quite well, until (roughly) the age of retirement. Consequently, we ran our model, and the visualisation, again, removing those over 64.

```{r pen lm & graph, cache = TRUE, message = FALSE, warning = FALSE}
W7p <- subset(W7, dvage < 65, select = c(1:8))

Model.6 <- lm(sclfsat1 ~ dvage, W7p)

summary(Model.6)

W7p %>%
  ggplot(aes(x = dvage, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE) +
  labs(title = "Figure 26",
       subtitle = "Trend in Wellbeing by Age (excluding pensioners)",
       x = "Age",
       y = "Wellbeing")
```

*Model 6* shows again that the relationship between age and wellbeing is statistically significant at the 99% confidence level (p<0.01). The effect of age on wellbeing is slightly larger than in *Model 5* (wellbeing decreases by 0.012 compared to 0.009 (4 S.F.) for each year older respondents are). *Figure 26* fits a linear model better than our last attempt. Interestingly, however, the graph shows a slight uptick in wellbeing just before the age of retirement, which, returning to *Figure 25*, continues for a few years before decreasing.

Having evaluated whether our individual variables had a statistically significant effect on wellbeing, we created a linear model where we controlled for each variable. *Model 7* shows that the significance of age, income, and sex are unchanged (all significant at the 99% confidence level, p<0.01), with largely similar effects. Interestingly, the difference between those who have higher degrees and those with AS-levels is now statistically significant at the 95% confidence level (p<0.05). AS-levels have a similar effect on wellbeing as A-levels. Additionally, the effect on wellbeing of higher education (e.g. ‘teaching qualification but not a PGCE’) has decreased, meaning there is now a noticeable difference between the effect of higher education and GCSEs. Finally, the effect of CSEs and ‘Other school certificates’ on wellbeing has decreased from -0.4185 to -0.3001 and from -0.5284 to -0.3154 respectively.

```{r all model, cache = TRUE, message = FALSE, warning = FALSE}
Model.7 <- lm(sclfsat1 ~ dvage + fimnnet + factor(qfhigh) + factor(sex), W7)

summary(Model.7)

# Note: these changes are for when all other variables are held constant, and represent the differences between respondents observed in our survey data. This does not, therefore, mean that if a respondent gained a new qualification, for example, their wellbeing would necessarily rise.
```

## Longitudinal analysis

So far, our analysis has focused on the most recent wave of data. However, here we first evaluate whether the overall changes in wellbeing are statistically significant *from year-to-year*, before doing this for each individual variable.

```{r wellbeing long, cache = TRUE, message = FALSE, warning = FALSE}
Model.8 <- lm(sclfsat1 ~ factor(year), d = dcontI)

summary(Model.8)
```

*Model 8* shows the difference in wellbeing from year-to-year, in comparison with 2009, is statistically significant at the 95% confidence level (p<0.05), and that the magnitude of this effect varies, though is fairly small in 2010 and 2015.

From our descriptive analysis section we can see that the wellbeing for men and women seems to follow the same pattern *Figure 4*, with no real change in the difference in wellbeing between the sexes. To evaluate this statistically, we used interaction effects between year and sex in a linear regression model, and found that the difference in wellbeing between the sexes is constant (there is no statistically significant difference at the 95% confidence level, p>0.05) (*Model 9*).

```{r sex long, cache = TRUE, message = FALSE, warning = FALSE}
Model.9 <- lm(sclfsat1 ~ as.factor(year) * sex, data = dcontI)

summary(Model.9)
```

For longitudinal analysis of education, we treated it as a time-constant variable, as all respondents are over the age of 18 when they respond to the survey and only a small number will gain a new qualification within the time-frame of our survey.

```{r education long, cache = TRUE, message = FALSE, warning = FALSE}
# We first installed the plm package.

library(plm)

Model.10 <- lm(sclfsat1 ~ as.factor(year) * factor(qfhigh), data = dcontI)

summary(Model.10)
```

Our analysis shows that, for the most part, the difference in wellbeing between respondents with different educational qualifications is constant (there is no statistically significant difference at the 95% confidence level, p>0.05) (*Model 10*). However, there are two exceptions to this. In 2014, those with ‘CSEs’ had a wellbeing that was lower than expected by 0.18 (3 S.F.), significant at a 95% confidence level (p<0.05) and in 2015, respondents with ‘Other school certificates’ had a wellbeing that was lower by 0.20 (3 S.F.), significant at a 95% confidence level (p<0.05).

So far we have seen a negative relationship between age and wellbeing (in 2015). However, as age is a time-varying variable, we can evaluate whether wellbeing changes when individuals age. *Model 11* shows that any changes in wellbeing as the participants in the survey aged were not statistically significant at the 95% confidence level (p>0.05).

```{r age long, cache = TRUE, message = FALSE, warning = FALSE}
Model.11 <- plm(sclfsat1 ~ dvage, data = dcontI, model = "within", index = c("pidp", "year"), effect = "twoways")

summary(Model.11)
```

Finally, we can evaluate whether changes in income for individuals have a statistically significant impact on wellbeing.

```{r income long, cache = TRUE, message = FALSE, warning = FALSE}
Model.12 <- plm(sclfsat1 ~ fimnnet, data = dcontI, model = "within", index = c("pidp", "year"), effect = "twoways")

summary(Model.12)
```

Surprisingly, this is not the case (this relationship is not significant at the 95% confidence level, p>0.05) (*Model 12*). Perhaps this is because a large increase in income is necessary for an impact to be made on an individual’s life (for example, so they can move to a nicer area) therefore any change in wellbeing due to an increase in income may take longer to actualise than we are able to see in our seven years of data.

# Discussion

Having established that changes in the general level of wellbeing across Britain were statistically significant from year-to-year, it is worthwhile looking at major events that could have shaped, or at least been indicative of, these changes.

* **November 2009:** Britain emerges from the longest recession since quarterly figures were first recorded (1955). The drop in wellbeing in the years following this could be the result of the recession and subsequent policies impacting on individuals. This could explain why the bottom 20% of earners seemed to be least affected by the general slump in average wellbeing, having perhaps the fewest savings to lose as a result of the recession.

* **May 2010:** The first post-war coalition government is formed. With a large proportion of the vote going to the Liberal Democrats, this may have indicated that the population were disenchanted with the established political parties, as they were unhappy with their lives.

* **October 2010:** Austerity began, with large-scale public spending cuts, averaging 19% over four years. General levels of wellbeing would have been expected to fall following this.

* **Over the course of 2011:** The UK became the world’s most indebted country and household income reached a low of 12th globally. Again, there is little wonder that general wellbeing fell following this.

* **August 2011:** There were widespread riots and looting in London and other major English cities after a young black man was shot and killed by a police officer. These riots may have been a manifestation of lower wellbeing, and could also have contributed to a fall in wellbeing as people became more concerned for their own safety.

* **May-June 2012:** The Queen’s Diamond Jubilee celebrations were a time of huge nationalistic pride across the UK.

* **August-September 2012:** British Olympics and Paralympics, which received “widespread international acclaim”, and were “seen as a fillip in a time of economic gloom”. This, coupled with the Jubilee celebrations, may have prevented general wellbeing from falling any further as a feeling of unity encompassed the UK.

* **October 2013:** Unemployment dropped below 2.5 million for the first time in 4 years. A rise in wellbeing following this is unsurprising.

* **By the end of 2014:** UK GDP growth was the fastest in Europe and the G7, with employment the highest since records began. Again, there is little wonder that general wellbeing increased following this.

* **May 2014:** There was a surge in UKIP votes in the election for the European Parliament.

* **September 2014:** The Scottish vote to remain as part of the UK.

* **May 2015:** The General Election resulted in the first Conservative majority since 1992 to the shock of the pollsters, with UKIP again making huge gains. Prior to 2016’s vote on the  European Union, and following the surge in UKIP votes and the Scottish referendum, nationalistic pride is perhaps at a high.

Sociological theory can perhaps explain the specific differences in wellbeing discovered within our variables. The existence of class-based differences in perceptions of wellbeing is partly explained by Bourdieu (1979), who argues that the working-classes possess a far greater ‘functional’ idea of health in comparison to the middle- and upper-classes. This difference suggests the working-classes are more likely to consider themselves as enjoying wellbeing as long as they can work, whereas the middle- and upper-classes equate wellbeing not only with being able to work, but also with certain aesthetic and fitness standards (for example, not being overweight and being able to participate in high status leisure activities with others) (Bourdieu, 1979; Shilling, 2012). This perhaps explains why the bottom 20% of earners, in particular those that are men, represent such an outlier in terms of wellbeing.

The differences in wellbeing between the sexes can also be explained, to an extent, by Bourdieu’s work. As men are under pressure to fulfil hegemonic ideals of masculinity (Connell, 1993), they may be likely to respond positively when asked how they are, regardless of the truth. Men are, for instance, notoriously poor at recognising and seeking help when they have mental or physical health issues. Indeed, this pressure to present oneself as ‘okay’ in order to meet dominant standards of masculinity could significantly have affected our results, and exacerbated the general issues with face-to-face surveys, as people often lie to fulfil social expectations (for example, saying they have voted when asked, whether they did or not). Again, this is differentiated by class, with Bourdieu (1979) suggesting that working-class men are under the greatest pressure to fulfil these ideals, which perhaps explains our unexpected results in terms of income.

From our longitudinal analysis we can conclude that the influence of sex, age, income and education have had a consistent effect on wellbeing over the period covered by our survey data. This conclusion can be used to argue against Giddens’s (1991, 1999) suggestion that the accelerated change and individualisation typical of late modernity has decreased the importance of traditional markers of social background on individuals’ lives and, specifically, wellbeing.

In terms of the impact of our study, we are most notably filling a gap in research on the effect education has on wellbeing, as it is an area of study that “lags behind” others in terms of its impact on wellbeing (Pinxten and Lievens, 2014: 1096). As we found that there is a positive correlation between education and wellbeing, the emphasis on encouraging and facilitating the education of individuals arguably takes on even greater importance. Indeed, this supports Abel’s (2007: 43) conclusion that “health promotion practice should give increased attention” to educating individuals about health “at the early stages of intervention planning”.


# Conclusion

Our investigation found that wellbeing decreased across the population from 2011-2013, before recovering in 2014 & 2015. We found that there was a statistically significant difference in wellbeing between the sexes, with men reporting higher wellbeing than women, contrary to our hypothesis. This difference did not change over time, as expected. We also found that older respondents reported lower wellbeing, though wellbeing was higher for those near the age of retirement. However, our longitudinal analysis showed that across the waves of the survey, no statistically significant change in wellbeing was observed as individuals aged. There were significant differences in wellbeing between education levels in our dataset, which were again mostly stable over time, as hypothesised. Finally, we found that income of £800 a month and over was positively correlated with wellbeing (until roughly £8,000 a month). No statistically significant differences in wellbeing were found for changes in individual income over time, against our hypothesis. Further investigation into the relationship between income and wellbeing would be particularly useful.

# Reference List

Abel, T. (2007) ‘Cultural capital in health promotion’, in D. McQueen, I. Kickbusch, L. Potvin, J.M. Pelikan, L. Balbo and T. Abel (eds), *Health and Modernity*. New York: Springer.

Bennett, T., Savage, M., Silva, E., Warde, A., Modesto, G-C. and Wright, D. (2009) *Culture, Class, Distinction*. London: Routledge.

Bourdieu, P. (1979) *Distinction: A Social Critique of the Judgement of Taste*. London: Routledge.

Connell, R.W. (1993) *Masculinities*. Cambridge: Polity Press.

Giddens, A. (1991) *Modernity and Self-Identity*. Stanford, CA: Stanford University Press.

Giddens A. (1999) ‘Risk and responsibility’, *The Modern Law Review*, 62(1): 1–10.

Grineski, S.E. (2009) ‘Parental accounts of children’s asthma care: the role of cultural and social capital in health disparities’, *Sociological Focus*, 42(2): 107–32.

Mackenbach, J.P. (2012) ‘The persistence of health inequalities in modern welfare states: the explanation of a paradox’, *Social Science & Medicine*, 75(4): 761–9.

Pinxten, W. and Lievens, J. (2014) 'The importance of economic, social and cultural capital in understanding health inequalities: using a Bourdieu-based approach in research on physical and mental health perceptions', *Sociology of Health and Illness*, 36(7): 1095-1110.

Samaritans, (2017) 'Suicide: Facts and Figures', https://www.samaritans.org/about-us/our-research/facts-and-figures-about-suicide. Accessed 26 April.

Shilling, C. (2012) *The Body and Social Theory*, 3rd Edition. London: Sage.

University of Essex. Institute for Social and Economic Research, NatCen Social Research, 	Kantar Public (2017) Understanding Society: Waves 1-7, 2009-2016 and Harmonised	BHPS: Waves 1-18, 1991-2009, [data collection], UK Data Service, 9th Edition,	Accessed 06 April 2018. SN: 6614, http://doi.org/10.5255/UKDA-SN-6614-10.