---
title: "Untitled"
author: '660032036'
date: "18 April 2018"
output: html_document
---

```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Health is an enormously important topic in sociology; for example, unequal morbidity and mortality rates are key indicators of social inequalities. However, while the differentiation between classes has been shown to be one of the “most enduring and incontrovertible indicators” of perceptions of personal health (referred to as wellbeing from this point on), it has been argued that more research is required in order to fully understand the dynamics of the social divides in wellbeing (Bennet et al., 2009: 152; Mackenback, 2012; Pinxten and Lievens, 2014). This essay therefore aims to further current research into wellbeing by taking into account sex, education and income: factors that may well contribute to wellbeing. As an individual’s life is separated into two main parts, their time in education and their time working, it makes sense to analyse the effects of both of these variables on individual wellbeing. Additionally, with male suicide rates over three times as high as female suicide rates, the difference in wellbeing between the sexes is as important as ever. [Samaritans site]

Our research question is ‘How is wellbeing distributed across the sexes, and what is the effect of different levels of education and income on wellbeing?’.

To carry out this analysis, we will be using the Understanding Society dataset, a longitudinal study of approximately 40,000 households in the United Kingdom, carried out from 2009-2016 and of a nationally representative sample. Each participant aged 16 and over completed an individual, face-to-face interview and a self-completed questionnaire, with one person in each household completing the household questionnaire.

Over this time period, there were a number of events that could have had a general effect on wellbeing across the nation. The UK entering and emerging from an economic recession, two general elections, and the 2012 Olympics were all significant events that need to be considered when analysing our data.

# Preparing the data

Our first task is to read the data into R, which is done by using the following code, after installing the necessary packages:

```{r loading, cache = TRUE, message = FALSE, warning = FALSE, results = 'hide'}
library(tidyverse)

library(data.table)

files <- dir("data/UKDA-6614-tab/tab",
             pattern="indresp",
             recursive = TRUE,
             full.names=TRUE)

files <- files[str_detect(files, "us")]

varsI <- c('pidp', 'qfhigh_dv', 'fimnnet_dv', 'fimngrs_dv', 'sclfsat1', 'dvage', 'sex', 'racel', 'racelo_code')

for (i in 1:7) {
  varsToSelectI <- paste(letters[i], varsI, sep = "_")
  varsToSelectI <- c("pidp", varsToSelectI)
  dataI <- fread(files[i], select = varsToSelectI)
  if (i == 1) {
    dI <- dataI  
  }
  else {
    dI <- full_join(dI, dataI, by = "pidp")
  }
  rm(dataI)
}
```

To arrange the data in a way so as to facilitate analysis, the ‘wave’ the data was collected in needs to be recoded as a variable:

```{r join, cache = TRUE, message = FALSE, warning = FALSE}
dmeltI <- dI %>%
  melt(id = "pidp") %>%
  separate(variable, into = c("wave", "variable"), sep = "_") %>%
  dcast(pidp + wave ~ variable)
```

The next step in preparing our data was removing the values for each variable that represent, for example, values that show the respondent refused to answer the question, and changing these to ‘NA’ values:

```{r NA, cache = TRUE, message = FALSE, warning = FALSE}
dnaI <- dmeltI %>%
  mutate(qfhigh = ifelse(qfhigh > 0, qfhigh, NA)) %>%
  mutate(qfhigh = ifelse(qfhigh < 96, qfhigh, NA)) %>%
  mutate(sclfsat1 = ifelse(sclfsat1 > 0, sclfsat1, NA)) %>%
  mutate(dvage = ifelse(dvage > 0, dvage, NA)) %>%
  mutate(sex = ifelse(sex > 0, sex, NA)) %>%
  mutate(racel = ifelse(racel > 0, racel, NA)) %>%
  mutate(racel = ifelse(racel < 96, racel, NA)) %>%
  mutate(racelo = ifelse(racelo > 0, racelo, NA)) %>%
  mutate(racelo = ifelse(racelo < 96, racelo, NA))
```

Initially, our research aimed to include ethnicity as a factor in our analysis. However, when checking our dataset, we realised that both variables for ethnicity had vast numbers of ‘NA’ values in each wave, with the ethnicity of the participant coded seemingly randomly across the waves of the data. Due to the limitations of this assignment, we decided simply to remove this variable. This process is shown below:

```{r race, cache = TRUE, message = FALSE, warning = FALSE}
# Subsetting the data frame into individual waves
W1 <- subset(dnaI, wave == 'a',
             select = c(1:10))

W7 <- subset(dnaI, wave == 'g',
             select = c(1:10))

# Checking for variables with larger than average number of NAs
summary(W1)

summary(W7)

dnaI <- subset(dnaI, , -c(racel, racelo, fimngrs))
```

The final step in preparing our data involved, firstly, recoding the answers given as numerical values to descriptive labels. We also combined equivalent qualifications, reducing the number of variables. These changes made our descriptive analyses both easier to carry out, and the results easier to interpret.

```{r clean, cache = TRUE, message = FALSE, warning = FALSE}
dcleanI <- na.omit(dnaI) %>%
  mutate(qfhigh = recode(qfhigh, "1" = "1-Higher degree", "2" = "2-Degree/equiv", "3" = "3-Higher ed",
                         "4" = "3-Higher ed", "5" = "3-Higher ed", "6" = "3-Higher ed",
                         "7" = "4-A level/equiv", "8" = "4-A level/equiv", "9" = "4-A level/equiv",
                         "10" = "5-AS level/equiv", "11" = "5-AS level/equiv", "12" = "5-AS level/equiv",
                         "13" = "6-GCSE/equiv", "14" = "7-CSE", "15" = "6-GCSE/equiv", "16" = "8-Other school cert")) %>%
  mutate(qfhigh = factor(qfhigh)) %>%
  mutate(sclfsat1 = recode(sclfsat1, "1" = "1-completely dissatisfied", "2" = "2-mostly dissatisfied",
                           "3" = "3-somewhat dissatisfied", "4" = "4-neither satisfied or dissatisfied",
                           "5" = "5-somewhat satisfied", "6" = "6-mostly satisfied", "7" = "7-completely satisfied")) %>%
  mutate(sclfsat1 = factor(sclfsat1)) %>%
  mutate(sex = recode(sex, "1" = "male", "2" = "female")) %>%
  mutate(sex = factor(sex)) %>% 
  mutate(year = recode(wave, "a" = "2009",
                       "b" = "2010",
                       "c" = "2011",
                       "d" = "2012",
                       "e" = "2013",
                       "f" = "2014",
                       "g" = "2015")) %>%
  mutate(year = as.numeric(year))
```

Secondly, we recorded our initial data to include the responses to wellbeing as numerical values, rather than the labels used above, in order to treat wellbeing as a continuous variable. This was so we could carry out longitudinal analysis of the effects of our variables on wellbeing.

```{r cont, cache = TRUE, message = FALSE, warning = FALSE}
dcontI <- na.omit(dnaI) %>%
  mutate(qfhigh = recode(qfhigh, "1" = "1-Higher degree", "2" = "2-Degree/equiv", "3" = "3-Higher ed",
                         "4" = "3-Higher ed", "5" = "3-Higher ed", "6" = "3-Higher ed",
                         "7" = "4-A level/equiv", "8" = "4-A level/equiv", "9" = "4-A level/equiv",
                         "10" = "5-AS level/equiv", "11" = "5-AS level/equiv", "12" = "5-AS level/equiv",
                         "13" = "6-GCSE/equiv", "14" = "7-CSE", "15" = "6-GCSE/equiv", "16" = "8-Other school cert")) %>%
  mutate(qfhigh = factor(qfhigh)) %>%
  mutate(sex = recode(sex, "1" = "male", "2" = "female")) %>%
  mutate(sex = factor(sex)) %>% 
  mutate(year = recode(wave, "a" = "2009",
                       "b" = "2010",
                       "c" = "2011",
                       "d" = "2012",
                       "e" = "2013",
                       "f" = "2014",
                       "g" = "2015")) %>%
  mutate(year = as.numeric(year))
```

Finally, before visualising the relationships between our variables and wellbeing, we removed the data files we no longer needed, as they were needlessly taking up memory.

```{r rm, cache = TRUE, message = FALSE, warning = FALSE}
rm(dI, dmeltI, dnaI)
```

# Descriptive analysis

### Distribution of wellbeing

First, we looked at the distribution of wellbeing across each wave of the data, trying to find any overall trends. We computed the mean average wellbeing score in each wave and visualised this, which showed us a noticeable drop in average wellbeing from 2010 to 2011, which (despite a small increase in 2012, perhaps due to the Olympics) stayed low until increasing again in 2014.

```{r wellbeing by year, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
 group_by(year) %>%
 summarise(
   meanWb = mean(sclfsat1, na.rm = TRUE)
 ) %>%
 ggplot(aes(x = year, y = meanWb)) +
 geom_point() +
 geom_line()
```

For a more detailed look at the trend in wellbeing over time, we looked at the proportion of responses for each level of wellbeing in each year of our data. The number of people both ‘mostly’ and ‘completely’ dissatisfied with their wellbeing decreased dramatically from 2010 to 2011, and a large reduction in the number ‘mostly satisfied’ with their wellbeing was also noticeable. While the number of people either ‘mostly’ or ‘completely’ dissatisfied with their wellbeing slowly decreased over this period, there was only a slight increase in 2012 of those ‘mostly satisfied’ with their wellbeing (again, perhaps due to the Olympics?), and the number ‘completely satisfied’ with their wellbeing continued to decrease. However, the number of people ‘mostly’ and ‘completely’ satisfied with their wellbeing almost returned to 2010 levels in 2014, and continued to improve in 2015, with those ‘mostly’ and ‘completely’ dissatisfied with their health seeing a similar drop.

```{r wellbeing breakdown, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(dcleanI, aes(x = year, fill = sclfsat1)) +
  geom_bar(position = 'fill')
```

This second chart allows us a closer look at the differences in wellbeing across the first, fourth and final waves of the data.

```{r select few, cache = TRUE, message = FALSE, warning = FALSE}
Wc <- subset(dcleanI, wave == c('a', 'd', 'g'),
             select = c(1:7))

ggplot(Wc, aes(x = wave, fill = sclfsat1)) +
  geom_bar(position = 'fill')
```

### Differentiating by sex

In order to take a more detailed look at the distribution of wellbeing across society, we first differentiate our data by sex of the respondents. This will allow us to carry out our analysis while controlling for sex, which, as per our hypothesis, is expected to have an effect on wellbeing. First, we created a simple line graph, showing the mean average wellbeing score for each wave, shown below.

```{r sex time, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
  group_by(year, sex) %>%
  summarise(
    meanWb = mean(sclfsat1, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = sex, linetype = sex, shape = sex)) +
  geom_point() +
  geom_line()
```

This shows a clear and consistent difference in wellbeing between the sexes, as per our hypothesis. Interestingly, however, average wellbeing for men is higher than for women, going against our hypothesis that men will have a lower average wellbeing than women.

Next, for a quick overview of the differences between wellbeing of men and women over the entire dataset, we create a stacked bar-chart, split by sex. This shows that, overall, women are proportionally more likely to be ‘completely dissatisfied’ with their wellbeing than men, though have slightly higher levels of being ‘completely satisfied’ with their wellbeing than men.

```{r sex coxcomb, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(dcleanI, aes(x = sex, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  coord_polar(theta = 'y')
```

To have a closer look at this difference, we compared the individual responses to wellbeing given by respondents over each wave. Using violin plots (below), we can see that for 2009 and 2010, the proportion of women who were ‘completely satisfied’ with their wellbeing is in fact higher than the proportion of men who said the same, despite the reverse being true for the categories ‘mostly’ and ‘somewhat’ satisfied. This is only true for these two years though, as following this, men had a higher proportion of responses in the positive categories for wellbeing. For each year, there was a higher proportion of women who were ‘completely dissatisfied’ with their wellbeing than men, though the overall proportion of people ‘completely dissatisfied’ than ‘completely satisfied’ was much lower, as indicated below.

```{r sex split, cache = TRUE, message = FALSE, warning = FALSE}
fe <- subset(dcontI, sex == 'female')

m <- subset(dcontI, sex == 'male')

ggplot(fe, aes(x = wave, y = sclfsat1, fill = wave)) +
  geom_violin()

ggplot(m, aes(x = wave, y = sclfsat1, fill = wave)) +
  geom_violin()

table(fe$sclfsat1, fe$wave)

table(m$sclfsat1, m$wave)

# Note: while this is not a table of proportion, the ratio between values is clear to see, and the overall table is easier to interpret than one showing proportion.
```

Another useful insight these graphs give us is an idea of how wellbeing has changed for men and women over time, especially in relation to one another. Returning to our line-graph, we can see that while the overall trend is similar (a dip in wellbeing over 2011-2013), the difference in mean wellbeing between men and women is relatively stable. This difference is analysed in further detail later on.

### Differentiating by education

The next step in our analysis of the distribution of wellbeing across society was to differentiate our data by education. As in our analysis of sex and wellbeing, we first made a simple line graph showing the mean average wellbeing score for each wave, segregated by level of education.

```{r edu, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
 group_by(year, qfhigh) %>%
 summarise(
   meanWb = mean(sclfsat1, na.rm = TRUE)
 ) %>%
 ggplot(aes(x = year, y = meanWb, colour = qfhigh)) +
 geom_point() +
 geom_line()

head(dcontI)
```

This shows some clear differences between levels of education and wellbeing, though not entirely as we predicted. A- and AS-levels appear to be interchangeable when evaluating their effect on wellbeing, although in 2013 those with AS-levels had, proportionally, an unusually high wellbeing mean average (even higher than those with a degree or equivalent). Further, those with AS-Levels were the only group whose wellbeing did not increase from 2011 to 2012. Interestingly, respondents with higher education, such as a 'teaching qualification (excluding pgce)', have similar scores of wellbeing as those with just GCSE/equivalent qualifications; noticeably lower than those with AS- or A-levels, or university degrees. Finally, those with ‘Other school certificates’ have a higher mean average than those with ‘CSE’s, with the exception of 2015.

For a closer overall look between education levels, we created a stacked bar-chart showing the overall proportions for each education level for wellbeing in each wave. Interestingly, this shows that respondents with AS- and A-levels are more likely to be ‘completely satisfied’ with their wellbeing than those with university degrees or higher, a finding that was not picked up in the previous chart. For a clearer comparison, we added a coxcomb chart comparing respondents with higher degrees, A-levels, and CSEs.

```{r edu breakdown, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(dcleanI, aes(x = qfhigh, fill = sclfsat1)) +
  geom_bar(position = 'fill')

Wd <- subset(dcleanI, qfhigh == c('1-Higher degree', '4-A level/equiv', '7-CSE'),
             select = c(1:7))

ggplot(Wd, aes(x = qfhigh, fill = sclfsat1)) +
  geom_bar(position = 'fill') +
  coord_polar(theta = 'y')
```

While this analysis is more detailed than our initial line-graph, we were able to develop this further by analysing these three qualification levels by each wave of data. This shows us that during the overall slump in wellbeing from 2011-2013, the proportion of those who were ‘completely’ and ‘mostly’ dissatisfied with their wellbeing increased the most in respondents with the highest education levels. However, the inverse is not true for those ‘completely’ and ‘mostly’ satisfied with their wellbeing, as it is those with A-levels, not with higher degrees who were least effected; in fact, the proportion of those with A-levels who were ‘completely satisfied’ with their wellbeing actually increased from 2010-2011.

```{r edu few, cache = TRUE, message = FALSE, warning = FALSE}
High <- subset(dcleanI, qfhigh == '1-Higher degree',
               select = c(wave, sclfsat1))

A.l <- subset(dcleanI, qfhigh == '4-A level/equiv',
            select = c(wave, sclfsat1))

CSE <- subset(dcleanI, qfhigh == '7-CSE',
              select = c(wave, sclfsat1))

ggplot(High, aes(x = wave, fill = sclfsat1)) +
  geom_bar(position = 'fill')

ggplot(A.l, aes(x = wave, fill = sclfsat1)) +
  geom_bar(position = 'fill')

ggplot(CSE, aes(x = wave, fill = sclfsat1)) +
  geom_bar(position = 'fill')
```

Following our analysis of the split between the sexes and wellbeing, it would be remiss not to compare this divide in wellbeing when analysing education. This analysis shows that the split between the sexes is exaggerated by certain education levels. Those who are least well educated (CSEs) have substantially higher levels of wellbeing for men than for women, and men with GCSEs have similar levels of wellbeing to those with A-levels and higher education, whereas for women those with GCSEs and higher education have substantially lower levels of wellbeing than those with A- and AS- levels. Interestingly, men with AS-levels and higher university degrees have the highest levels of wellbeing in men, where women with higher university degrees and university degrees have the highest levels of wellbeing in women. The only qualification where women score higher on wellbeing than men is university degrees.

```{r edu sex, cache = TRUE, message = FALSE, warning = FALSE}
dcontI %>%
 group_by(year, qfhigh, sex) %>%
 summarise(
   meanWb = mean(sclfsat1, na.rm = TRUE)
 ) %>%
 ggplot(aes(x = year, y = meanWb, colour = qfhigh, linetype = sex, shape = sex)) +
 geom_point() +
 geom_line()
```

### Differentiating by income

The final part of our descriptive analysis brought us on to the topic of income. As we used the variable for net monthly income, we decided to recode this into quantiles at every 20% of income, to make our descriptive analysis easier both to do, and to interpret.

```{r income code, cache = TRUE, message = FALSE, warning = FALSE}
quantile(dcleanI$fimnnet, seq(0, 1, 0.2))

sc <- dcleanI %>% 
  mutate(income = ifelse(fimnnet >= -17741.3379 & fimnnet <= 518.9808, 20,
                         ifelse(fimnnet > 518.9808 & fimnnet <= 1051.6666, 40,
                                ifelse(fimnnet > 1051.6666 & fimnnet <= 1502.5000, 60,
                                       ifelse(fimnnet > 1502.5000 & fimnnet <= 2150.7295, 80,
                                              ifelse(fimnnet > 2150.7295 & fimnnet <= 15000.0000, 100, NA)))))) %>% 
  mutate(income = as.factor(income))
```

Our first graph is an overview of the dataset when segregating respondents by income. Incredibly, the group who are proportionately most likely to be ‘completely satisfied’ with their wellbeing are the bottom 20% of earners. This graph also shows that this group also have the lowest rates of dissatisfaction with wellbeing, and by looking at a table of this data for a more precise look, we can see the bottom 20% of earners are the third least likely to be ‘completely dissatisfied’, second least likely to be ‘mostly dissatisfied’ and least likely to be ‘somewhat dissatisfied’ in terms of wellbeing. This, coupled with the fact that the 4 other income brackets have almost identical proportions of respondents who were ‘completely satisfied’ with their wellbeing, is entirely contrary to our hypothesis that income is positively correlated with wellbeing.

```{r income overview, cache = TRUE, message = FALSE, warning = FALSE}
ggplot(sc, aes(x = income, fill = sclfsat1)) +
  geom_bar(position = 'fill')

table(sc$sclfsat1, sc$income)
```

Our first line-graph shows almost exactly what we hypothesised: as income decreases, so does the mean average wellbeing. However, there is one very substantial exception to this rule. Those in the bottom fifth of earners have exceptionally high mean average wellbeing levels, not far from the top fifth of earners. This is a very surprising result, however could be quite easy to explain. Notably, net monthly income can be negative. As this variable is purely personal income, it does not take into account other earners in the household. This could mean that the household net income is in fact in the top 20%, yet this is earnt by only one person in the household, meaning other family members benefit from this secondary wealth. To try to counter this possible effect, we then removed individuals who had a negative net income, as we theorised that those on low-salaries would not spend far outside their means, unlike those with a spouse or parent on a high income salary.

```{r income t & s, cache = TRUE, message = FALSE, warning = FALSE}
quantile(dcontI$fimnnet, seq(0, 1, 0.2))

t <- dcontI %>% 
  mutate(income = ifelse(fimnnet >= -17741.3379 & fimnnet <= 518.9808, 20,
                         ifelse(fimnnet > 518.9808 & fimnnet <= 1051.6666, 40,
                                ifelse(fimnnet > 1051.6666 & fimnnet <= 1502.5000, 60,
                                       ifelse(fimnnet > 1502.5000 & fimnnet <= 2150.7295, 80,
                                              ifelse(fimnnet > 2150.7295 & fimnnet <= 15000.0000, 100, NA)))))) %>% 
  mutate(income = as.factor(income))

t %>%
  group_by(year, income) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income)) +
  geom_point() +
  geom_line()

quantile(dcontI$fimnnet[dcontI$fimnnet > 0], seq(0, 1, 0.2))

s <- dcontI %>% 
  filter(fimnnet > 0) %>% 
  mutate(income = ifelse(fimnnet >= 1.506607e-04 & fimnnet <= 6.706440e+02, 20,
                         ifelse(fimnnet > 6.706440e+02 & fimnnet <= 1.125058e+03, 40,
                                ifelse(fimnnet > 1.125058e+03 & fimnnet <= 1.567589e+03, 60,
                                       ifelse(fimnnet > 1.567589e+03 & fimnnet <= 2.200000e+03, 80,
                                              ifelse(fimnnet > 2.200000e+03 & fimnnet <= 1.500000e+04, 100, NA)))))) %>% 
  mutate(income = as.factor(income))

s %>%
  group_by(year, income) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income)) +
  geom_point() +
  geom_line()
```

This change in coding made had a substantial effect on the outcome, although the bottom fifth of earners still represent an outlier in mean average wellbeing. Interestingly, this group seemed to be least effected by the general slump in average wellbeing from 2011-2013, going from second bottom to second top (in mean average wellbeing) over this period. One possible explanation for the bottom fifth still being an outlier is retirees, whose net monthly income may not be very high, but whose savings may help them spend well. We correct for this by sub-setting our data to only include those under 65 (the retirement age in the UK), though this has very little difference overall, suggesting other factors are influential in this outlier.

```{r income y1, cache = TRUE, message = FALSE, warning = FALSE}
y <- subset(s, dvage < '65',
           select = c(1:9))

table(y$dvage)

y1 <- subset(y, dvage!= 100)

table(y1$dvage)

y1 %>%
  group_by(year, income) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income)) +
  geom_point() +
  geom_line()

# Note: for some reason, one respondent who was 100 was not removed by the initial sub-setting, hence having to subset the data a second time. 
```

The final part of our descriptive analysis involves splitting this final graph by sex, given the notable difference it made in wellbeing generally. Indeed, by segregating between sexes, this graph shows some remarkable trends. Much like for education, there is only one level of income that women have a higher mean average wellbeing than men, in this case the second fifth of respondents. For every other level of income, men have a higher mean average wellbeing each year, almost without exception. Arguably just as notably is the mean average wellbeing for the bottom fifth of earners when split by sex. For men, this group’s wellbeing hovers around the same level as those in the second highest bracket of earners, and nearly matches those in the top bracket of earners in the most recent wave of survey data. For women, the bottom 20% of earners are either similar to the lowest mean average level for wellbeing, or indeed score lowest. Furthermore, the difference between the male and female levels for wellbeing are most different in this category overall, by some distance, with only the sex divide in wellbeing for the top 20% of earners being comparable, in 2013.

```{r income y1 sex, cache = TRUE, message = FALSE, warning = FALSE}
y1 %>%
  group_by(year, income, sex) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income, linetype = sex)) +
  geom_point() +
  geom_line()
```

There are a couple of possible explanations for these disparities. First, women in the bottom 20% of earners are perhaps more likely to be housekeepers than men, having to work long and hard without pay. Men in this category are perhaps more likely to have a high gross salary with a large amount of outgoings; more data would be useful here to evaluate these theories. Second, men in the second fifth of lowest earners are perhaps likely to be working in jobs requiring heavy manual labour, with higher risk of personal injury, longer, more taxing hours, and fewer health benefits or insurances, whereas women in this category are perhaps more likely to be working in roles such as cleaning or catering. Further, the gap in average wellbeing between this second lowest category of earners and other levels is generally the largest for men. The same is not true for women, with the largest gap in average wellbeing coming between those best paid and the other levels.

Unfortunately, visualising the differences in wellbeing between respondents with different levels of income, education, and of different sexes is impractical, and this comparison will be left for statistical analysis (shown below is the graph split by each of these three factors. Aside from a couple of notable outliers, it is almost entirely uninterpretable).

```{r y1 all, cache = TRUE, message = FALSE, warning = FALSE}
y1 %>%
  group_by(year, income, sex, qfhigh) %>%
  summarise(
    meanWb = mean(sclfsat1)
  ) %>%
  ggplot(aes(x = year, y = meanWb, colour = income, linetype = sex, shape = qfhigh)) +
  geom_point() +
  geom_line()
```

# Statistical analysis

Having taken a preliminary look into our dataset, and visualised some of the relationships of interest, we move on to evaluating whether any of these relationships are statistically significant. First, we calculated whether there was a significant relationship between the age of respondents and their wellbeing (using the most recent wave of data).

```{r age lm, cache = TRUE, message = FALSE, warning = FALSE}
W7 <- subset(dcontI, wave == 'g', select = c(1:8))

age <- lm(sclfsat1 ~ dvage, W7)

summary(age)
```

This shows that, in our sample, the older the respondent, generally the lower score they gave for their wellbeing. The following graph gives a visual representation of this relationship.

```{r age graph, cache = TRUE, message = FALSE, warning = FALSE}
W7 %>%
  ggplot(aes(x = dvage, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE)
```

This graph shows that a linear model seems to fit the relationship between age and wellbeing quite well to an extent; the linear model fits less well at what looks to roughly be the age of retirement. Therefore, we run our model, and the visualisation of this, again, taking account of this observation.

```{r pen lm & graph, cache = TRUE, message = FALSE, warning = FALSE}
W7p <- subset(W7, dvage < 65, select = c(1:8))

agepen <- lm(sclfsat1 ~ dvage, W7p)

summary(agepen)

W7p %>%
  ggplot(aes(x = dvage, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE)
```

This model shows again, the relationship between age and wellbeing is statistically significant, and that the difference between people of different ages is slightly larger (wellbeing smaller by 0.011 compared to 0.009 for each year between respondents). The graph fits a linear model better than our last attempt, though not perfectly. Interestingly, the graph shows a slight uptick in wellbeing in the couple of years before the age of retirement, which, returning to our previous graph, continues for a few years afterwards then drops again. Whether this relationship remains statistically significant when controlling for the other variables we are analysing will be considered later.

Next, we compared the relationship between income and wellbeing. Again, using data from the most recent wave, we used a linear model to see if the relationship between income and wellbeing was statistically significant. We then visualised this relationship.

```{r income lm & graph, cache = TRUE, message = FALSE, warning = FALSE}
income <- lm(sclfsat1 ~ fimnnet, W7)

summary(income)

W7 %>%
  ggplot(aes(x = fimnnet, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE)
```

Our model shows there is a statistically significant relationship between income and wellbeing, though our visualisation of this relationship shows that our model may not be best suited to capturing this relationship. Indeed, as indicated in our descriptive analysis previously, the relationship between income and wellbeing is less clear for respondents with low, or negative, monthly income. In fact, our next graph shows this relationship become far clearer (and indeed, as our new model shows, has a larger effect (7.256e-05 compared to 3.951e-05)) for respondents who earn over £800 a month, equating to £9,600 a year. Interestingly, this is roughly the amount an individual on minimum wage, working a 40-hour week for 45 weeks (taking into account bank-holidays and standard holiday allowance) will earn after tax (shown below). It seems, therefore, that when in full time employment, wellbeing increases with income (until about £8,000 a month at which point the effect tails off), though for individuals earning less than this there are a number of other confounding factors.

```{r minimum income, cache = TRUE, message = FALSE, warning = FALSE}
W7r1 <- subset(W7, fimnnet > 800,
              select = c(1:8))

W7r1 %>%
  ggplot(aes(x = fimnnet, y = sclfsat1)) +
  geom_smooth() +
  geom_smooth(method = "lm", colour = "red", se = FALSE)

incomer1 <- lm(sclfsat1 ~ fimnnet, W7r1)
summary(incomer1)

6.7*40*45*.8

# Calculated using the minimum wage for over-21s (https://www.gov.uk/national-minimum-wage-rates) for 2015 (the final full working year before survey data was taken throughout 2016), subtracting holiday time (https://www.gov.uk/bank-holidays and https://www.gov.uk/holiday-entitlement-rights), after tax (https://www.gov.uk/income-tax-rates/previous-tax-years).
```

The final relationship we will look at before creating a model that controls for all our variables is that between education and wellbeing.